Probability of causal contacts between interstellar civilizations through Monte Carlo simulations

### 2020-05-25 submit_v03/manuscript.tex
advence

@@

We stress the fact that the only observation that can be stated with
certainty is that for the number of years projects for the Search of
Extraterrestrial Intelligences (SETI) have been working; we have
received no signal using the technical possibilities and the
conditions established by them \citep{tarter_search_2001}.

@@

the SETI initiatives, noticeable overshadowed because despite decades
of effort, not a single evidence for intelligent signals has been
detected.

@@

In a recent work, \citep{balbi_impact_2018} use a statistical model to
analyse the occurrence of causal contacts between civilisations in the
Galaxy.
%
The author highlights the effect of evolutionary processes when
attempting to estimate the number of communicating civilisations that
might be in causal contact with an observer on the Earth.
%
\citet{cirkovic_temporal_2004} also emphasise the lack of temporal
structure in the Drake equation.
%
\citet{bloetscher_using_2019} uses a probabilistic approach, though
still motivated by the Drake equation, to obtain a probabilistic
measure for the number of civilisations in the Galaxy.
%
To that end, the authors do Monte-Carlo Markov Chains over each factor
of the Drake equation, and combine the means to reach a probabilistic
result.
%
It is worth mentioning that the author proposes a log-normal target
distribution to compute the posterior probabilities.
%
This study concludes that there is a tiny probability for Earth to
contact another extraterrestrial civilization.    
%
\citet{smith_broadcasting_2009} uses an analytical model to gauge the
probabilities of contact between two randomly located civilisations,
and the waiting time for the first contact, assuming a fixed maximum
broadcasting distance.
% 
Adopting a novel approach, \citet{balbi_impact_2018} investigate the
chance of communicating civilisations making causal contact within a
volume surrounding the location of the Earth, using a statistical
model to compute the occurrence of causal contacts.
%
The author stresses the fact that the causal contact requirement
involves the distance between civilisations, their lifespan and their
times of appearance. This is important since the time it takes the
light to travel across the Galaxy is much lesser than its lifetime.
%
\citet{balbi_impact_2018} fix the total number of civilisations and
explore the parameter space that comprise three parameters, namely,
the distance to the Earth, the time of appearance and the lifespan of
the communicating civilization.
%
Each of these three variables are drawn from a random distribution.
The distribution of the distances results from a uniform distribution
of civilizations within the plane of the Galaxy.
%
For the distribution of the characteristic time of appearance, the
author explores an exponential and a truncated Gaussian distributions.
For the distribution of the lifespans, the author chooses an
exponential distribution.
%
\citet{balbi_impact_2018} finds that the fraction of communicating
civilisations vary with the choice of the statistical distribution for
the time of appearance. 
%
For all analysed distributions, the fraction corresponding to a mean
lifetime of 10$^9$ years and for a maximum radius for the detection of
the signal of 1000~ly, is roughly 0.1.
%
In another related approach, \citet{vukotic_astrobiological_2012}
propose a probabilistic cellular automata (PCA) modelling. 
%
In this framework, a complex system is modelled by a lattice of cells
which evolve at discrete time steps, according to transition rules
that take into account for each cell the states of its neighbour
cells.
%
The authors implement a PCA model to a network of cells which
represent life complexity on a 2-dimensional region resembling the
Galactic Habitable Zone (GHZ),  an annular ring set between a minimal
radius of 6 kpc and a peak radius of 10 kpc. 
%
Within this framework, \citet{vukotic_astrobiological_2012} also make
several Monte Carlo simulations and analyse ensemble-averaged results.
%
The work aims at analysing the evolution of life, although it does not
account for the network of causal contacts among
technological civilisations.

In this work we address the problem of the temporal and spatial
structure of the distribution of communicating civilisations.
%
To that end, we reduce and explore the hypothesis space over a set of
three of parameters.
%
We propose to avoid the frequentist approach of the Drake equation to
compute the number of civilisations, providing its statistical
distribution instead.
%
This is an exploratory analysis that aims at providing a numerical
tool to discuss the different scenarios based on statistical
heuristics.
%
With this, we aim at avoiding the several theoretical problems
summarised by the Drake equation factors.
%
The approach proposed here should be considered as a compromise
between the uncertainties of the frequentist approach and the detailed
recipes required on the simulation approach.
%
We provide a framework to explore, through a suite of
numerical simulations, a parameter space of three unknown observables.
%
This allows to discuss possible scenarios and their consequences in
terms of the probability of making contacts.
%
The method we use for simulating a stochastic process is an
approximation that allows to study the behaviour of complex systems,
by considering a sequence of well defined discrete events.
%
The simulation is carried out by following all the variables that
describe and constitute the state of the system, and the evolution of
the process is described as a set of changes in that state.
%
In this context, an event produces a specific change in the state that
can be triggered by random variables that encode the stochastic nature
of the physical phenomenon.
%
For example, when a new contact is produced between two entities in
the simulated galaxy, the numbers of active communication lines and of
communicated nodes does change.
%
The process involves following the changes on the state of the system,
defining the initial and final states.
%
This is done by defining a method that allows to keep track of the
time progress in steps, and maintaining a list of relevant events.
%
In Sec.~\ref{S_methods} we introduce the methods and discuss the
candidate distributions for the statistical aspects of the times
involved in the communication process.
%
We present our results in Sec~\ref{S_results}, with special emphasis
on the statistical distributions of the duration of causal contacts in
one or both directions and the distribution of time intervals for the
waiting of the first contact.
%
This quantities are considered as a function of the three simulation
parameters.
%
In Sec.~\ref{S_discussion} we discuss our results and future research
directions.   
%
In this work we use the International System of units recommended by
the International Astronomial Union, where the simbol for years is
''a''.
%}}}

@@

\ffn{1} Space—time diagrams, where time and space are represented on
the horizontal and vertical axes, respectively, are suitable tools to
describe the causal connections among different nodes.
%
We illustrate in the Fig.~\ref{F_scheme} the schematic representation
of the region causally connected to a given \ceti{}.
%
For the sake of simplicity, we show in the plot the light travel
distance, i.e., the distance traversed by any signal spreading from
the emitter at light speed (for example, any electromagnetic signal).
%
In this scheme, a light pulse would follow a trajectory represented by
a line at 45 degrees from the axes, given the units of the time and
space axes are years and light years, respectively.
%
Panel (a) of the Figure represents the region $R$ (shaded polygon)
which is causally connected to the emitter (left vertex).
%
This region develops after an event (hereafter dubbed A-type event) in
which the node acquires the communication capability, or becomes
''active''.
%
The sphere of first contact (SFC) is centred on the emitter,
represented in the Figure by the left angle of the region $R$.
%
This sphere grows until its radius reaches the D$_{\rm max}$ distance, in
which it would be no longer detectable due to the decrease in the
energy per unit area, which falls under the assumed detectability
threshold.
%
Similarly, the surface of last contact (SLC) is another sphere that
grows from an event in which the nodes ceases to possess the
communication capability (D type event), and carries the last signal
produced by the emitter.
%
The region which is causally connected to the emitter is then limited
by these two spheres and therefore has the shape of a filled sphere
(before the D event) or of a spherical shell (after the D event).
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b) of the Figure.
%
In this scheme, the entire lifetime of a \ceti{} can be represented by
a polygon, as shown in the Figure.
%
The polygon is limited in the spatial axis by the maximum distance of
the signal $D_{\rm max}$ and in the time axis by the time span between the
A-type event (concave vertex) and the D-type event (convex vertex).
%
This representation allows to visualize the important events that
result from the presence and communications of two \cetis{}.
%
The points in time where the \cetis{} acquire their communicating
capacity are dubbed ‘’Awakening’’ events ($A_1$ and $A_2$ in the
Figure).
%
Similarly, the instants in time where the \cetis{} lose their
communicating capacity are dubbed ‘’Doomsday’’ events ($D_1$ and $D_2$
in the Figure).
%
The points in space--time where the first contact is produced for each
one of the \cetis{}, are defined as ''Contact'' events, shown as $C_1$
and $C_2$.
%
Finally, the two points where the contact is lost for each one the
\cetis{}, are denominated ''Blackout'' events, presented as $B_1$ and
$B_2$ in the Figure.
%
The receiver node E$_1$ can listen signals from the emitter node
E$_2$, from the first contact event (t=C$_{12}$) until the last
one (t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$.
%
The time intervals for the open communication channel are determined
by the ''C'' and ''B'' type events.
%
It is noteworthy the fact that there is a time delay between the
contact events of the two \cetis{} involved in this analysis, and also
a time delay between the two blackout events.
%
Therefore, the time range when a bidirectional contact is possible
occurs between the maximum time of the contact points and the minimum
time between the blackout points.

\begin{figure*}
   \centering
   \includegraphics[width=\textwidth]{F_messages.pdf}
   \caption{
   %   
Schematic representation on space--time diagrams of the possible cases in which a
\ceti{} E (Emiter, solid line) can be in causal contact with a \ceti{} R
(Receiver, dashed line).
%
Causal contact can be produced either if the emiter \ceti{} appears before
(left) or after (right) the receiver \ceti{}.
%
The causal contact in which the receiver can have access to the signal 
from the emiter is produced between the C event (Contact, open circles) and the B
event (Blackout, filled circles).
%
The duration of the causal contact in one direction depends on
several factors, mainly the time lag between the awakening events
(horizontal separation in the space--time diagrams)
and the distance between the \cetis (vertical separation in the
space--time diagrams).
%   
The different configurations must be taken into account in order to 
obtain the list of the upcomming events.
%
The time intervals for a two--way communication to be possible are
indicated in double lines. 
%
   } \label{F_messages}
\end{figure*}

The temporal structure that emerges from the experimental setup implies that, as a fundamental property of the simulations, a causal connection can be produced without 
requiring that the two nodes are active at the same time.
This property of the system arises as a consequence of the large spatial and temporal scales, where a message is transmitted at the (relatively small) light speed.
Although a \ceti{} could be active for a large enough period to transmit a message at large distances, the limited power of the message and the dilution that depends on the squared distance from the source imposes a detectability limit.
As a consequence of this limitation and of their finite lifetime, considered as the period between the acquisition and loss of communicating capacity, each \ceti{} will fill a spherical shell region of the galaxy, limited by two concentric spherical surfaces.
The leading front, or surface of first contact (SFC) grows from the
central \ceti{} until it reaches the maximum distance $D_{\rm max}$.
Following the end of the civilization, there is still a region which is filled with the emitted signals.
This approach has been also considered in other statistical models \citep[e.g., ][]{smith_broadcasting_2009, grimaldi_signal_2017, Grimaldi2018}.
The trailing front, or surface of last contact (SLC) also grows from the central \ceti{}, with a delay with respect to the SFC equivalent to the lifetime of the node, and produces a spherical shell region.
Any other \ceti{} within this region will be aware of the originating node, even if it has disappeared before the time of contact.
This region will grow if the surface of first contact has not yet
reached the maximum distance $D_{\rm max}$, and will shrink after a
D-type event until the surface of last contact reaches $D_{\rm max}$, producing as a result the loss of all signal from the central \ceti{}.
In our approach, we consider a model galaxy where the width of the disk is negligible with respect to the radius of the disk.
In the 2D simulation only the intersection of the communicating spherical shells with the plane of the galaxy is relevant, and produce the corresponding circles or rings for the filled spheres or annular regions of the spherical shells, respectively.
The initially growing communicating sphere is shown over space--time diagrams, where time is represented on the horizontal direction, and space is represented in the vertical direction.
In the Fig.~\ref{F_scheme} the two emitters in panel (b), E$_1$ and E$_2$, reach each other at different times.
The time span for E$_i$ is (A$_i$ , D$_i$) , for $i = {1, 2}$.
Emitter $i$ can listen to emiter $j$ between C$_{ij}$ and B$_{ij}$.
The type and length of causal contact in both directions depend on the distance and time lag between the awakening events, the maximum distance that a signal can reach and the time period in which each emitter is active.
\ffn{2}
In the Fig.~\ref{F_messages} we show a schematic representation of the possible cases in which a \ceti{} E (standing for Emiter) can be in causal contact with a \ceti{} R (receiver).
The duration of such a contact in one way depends on several factors, mainly the lifetimes, the time delay between the A-type events and the distance between the two nodes.
As it can be seen, the direction of the causal contact does not imply any ordering for the A-type events of the involved emitter/receiver pair.
For instance, a node can be in the light cone of another node that turned up later, i.e., the receiver is older than the emitter, like in the cases (e) and (f) of the Figure.
In the implementation of the DE simulation, all possibilities for causal contacts are reduced to the six cases shown in the Figure.
Once a pair of nodes become active and their properties have been assigned, it is possible to determine if and when they will get connected.
The times derived for such events are added to a stack of times so that they are visited in order, and in turn trigger further events.
% EJEMPLOS???

@@

 \setlength{\tabcolsep}{10pt}
\begin{table*}
\centering
\begin{tabular}{cllll}
\hline
   \multicolumn{2}{l}{independent variables (hypothesis space)}
   &min. value&max. value&Nbins\\
\hline
   $\tau_{a}$ & Mean temporal separation between consecutive
   awakenings & 10$^2$~a & 10$^6$~a & 50\\ 
   $\tau_{s}$ & Mean lifetime of a \ceti{}
   & 10$^2$a & 10$^6$~a& 50\\ 
   D$_{\rm max}$ & Maximum reach of a message  & 30~pc & 10~kpc & 50 \\
\hline
   \multicolumn{2}{l}{fixed variables} & \multicolumn{2}{c}{assumptions} &value \\
\hline
   & statistical properties of all \cetis{} &equally distributed&&\\
   & Point process for the distribution in time & homogeneous &&\\
   f$_s$ & The scan of the sky & fully efficient&&1\\
   f$_p$ & panspermia or colonization &absent&&0\\
   & shape of the Galactic Habitable Zone & 2--dimensional ring &&\\
   R$_{GHZ}^{\rm min}$   & Inner radius of the GHZ     & & & 7~kpc\\
   R$_{GHZ}^{\rm max}$   & Outer radius of the GHZ     &&& 9~kpc\\
   t$_{\rm max}$ & Time span of the simulation  &&&1.e6~a\\
    & \multicolumn{3}{l}{number of random realizations for each point in the hypothesis space} &50\\
\hline
   \multicolumn{2}{l}{discrete events} &
   \multicolumn{3}{l}{affected variables}\\
\hline
   A event & Awakening: a \ceti{} starts its communication
   capabilities &Number of active \cetis{}\\
   B event & Blackout: the end of the communication channel stops
   &Number causal channels\\
   C event & Contact: a new causal contact is produced &Number of causal channels\\
   D event & Doomsday: a \ceti{} ends its communication capabilities&Number of active \cetis{}\\
\hline

>>>>

We stress the fact that the lack of detections of extraterrestrial intelligent signals
is
the only observational fact that can be taken
for sure.

@@

Indeed, for the number of years projects for the Search of
Extraterrestrial Intelligences (SETI) have been working,
no signal has been received using the technical possibilities 
within their search criteria \citep{tarter_search_2001}.

@@

the results from the SETI initiatives.
Notably, despite decades of effort not a single evidence for intelligent signals has been detected.

@@

In a recent work, \citep{balbi_impact_2018} use a statistical model to
analyse the occurrence of causal contacts between civilisations in the
Galaxy.
%
The author highlights the effect of evolutionary processes when
attempting to estimate the number of communicating civilisations that
might be in causal contact with an observer on the Earth.
%
\citet{cirkovic_temporal_2004} also emphasise the lack of temporal
structure in the Drake equation.
%
\citet{bloetscher_using_2019} uses a probabilistic approach, though
still motivated by the Drake equation, to obtain a probabilistic
measure for the number of civilisations in the Galaxy.
%
To that end, the authors do Monte-Carlo Markov Chains over each factor
of the Drake equation, and combine the means to reach a probabilistic
result.
%
It is worth mentioning that the author proposes a log-normal target
distribution to compute the posterior probabilities.
%
This study concludes that there is a tiny probability for Earth to
contact another extraterrestrial civilization.    
%
\citet{smith_broadcasting_2009} uses an analytical model to gauge the
probabilities of contact between two randomly located civilisations,
and the waiting time for the first contact, assuming a fixed maximum
broadcasting distance.
% 
Adopting a novel approach, \citet{balbi_impact_2018} investigate the
chance of communicating civilisations making causal contact within a
volume surrounding the location of the Earth, using a statistical
model to compute the occurrence of causal contacts.
%
The author stresses the fact that the causal contact requirement
involves the distance between civilisations, their lifespan and their
times of appearance. This is important since the time it takes the
light to travel across the Galaxy is much lesser than its lifetime.
%
\citet{balbi_impact_2018} fix the total number of civilisations and
explore the parametric space that comprises three parameters, namely,
the distance to the Earth, the time of appearance and the lifespan of
the communicating civilization.
%
Each of these three variables are drawn from a random distribution.
%
The distribution of the distances results from a uniform model for the
positions of civilizations within the plane of the Galaxy.
%
For the distribution of the characteristic time of appearance, the
author explores exponential and truncated Gaussian distributions.
%
Moreover, for the distribution of the lifespans, the author chooses an
exponential distribution.
%
\citet{balbi_impact_2018} finds that the fraction of communicating
civilisations vary with the choice of the statistical distribution for
the time of appearance. 
%
For all analysed distributions, the fraction corresponding to a mean
lifetime of 10$^9$ years and for a maximum radius for the detection of
the signal of 1000~ly, is roughly 0.1.
%
In another related approach, \citet{vukotic_astrobiological_2012}
propose a probabilistic cellular automata (PCA) modelling. 
%
In this framework, a complex system is modelled by a lattice of cells
which evolve at discrete time steps, according to transition rules
that take into account for each cell the states of its neighbour
cells.
%
The authors implement a PCA model to a network of cells which
represent life complexity on a 2-dimensional region resembling the
Galactic Habitable Zone (GHZ),  an annular ring set between a minimal
radius of 6 kpc and a peak radius of 10 kpc. 
%
Within this framework, \citet{vukotic_astrobiological_2012} also make
several Monte Carlo simulations and analyse ensemble-averaged results.
%
The work aims at analysing the evolution of life, although it does not
account for the network of causal contacts among
technological civilisations.

% INTRODUCTION OF *THIS* WORK

In this work we address the problem of the temporal and spatial
structure of the distribution of communicating civilisations.
%
We make no asumptions about the origin of life, the development 
of intelligence or the formation of habitable planets according to stellar type.
%
Instead, we assume monoparametric function families to model the
appeareance of points in the disk of the Galaxy

%
To that end, we propose a model which depends on a set of only
three parameters.
%
We abandon the frequentist approach of the Drake equation to
compute the number of civilisations, providing its statistical
distribution instead.
%
This is an exploratory analysis that aims at developing a numerical
tool to discuss the different scenarios based on statistical
heuristics.
%
With this, we avoid the several theoretical problems
summarised by the Drake equation factors.
%
The approach proposed here should be considered as a compromise
between the uncertainties of the frequentist estimations and the detailed
recipes required on the numerical simulations.
%
We provide a framework to explore, through a suite of
numerical simulations, a parameter space of three unknown observables.
%
This allows to discuss possible scenarios and their consequences in
terms of the probability of making contacts.
%
The method we use for simulating a stochastic process is an
approximation that allows to study the behaviour of complex systems,
by considering a sequence of well defined discrete events.
%
The simulation is carried out by following all the variables that
describe and constitute the state of the system, and the evolution of
the process is described as a set of changes in that state.
%
In this context, an event produces a specific change in the state that
can be triggered by random variables that encode the stochastic nature
of the physical phenomenon.
%
For example, when a new contact is produced between two entities in
the simulated galaxy, the numbers of active communication lines and of
communicated nodes does change.
%
The process involves following the changes on the state of the system,
defining the initial and final states.
%
This is done by defining a method that allows to keep track of the
time progress in steps, and maintaining a list of relevant events.
%
In Sec.~\ref{S_methods} we introduce the methods and discuss the
candidate distributions for the statistical aspects of the times
involved in the communication process.
%
We present our results in Sec~\ref{S_results}, with special emphasis
on the statistical distributions of the duration of causal contacts in
one or both directions and the distribution of time intervals for the
waiting of the first contact.
%
This quantities are considered as a function of the three simulation
parameters.
%
In Sec.~\ref{S_discussion} we discuss our results and future research
directions.   
%
In this work we use the International System of units recommended by
the International Astronomial Union, where the simbol for years is
''a''.
%}}}

@@

\ffn{1} Space—time diagrams, where time and space are represented on
the horizontal and vertical axes, respectively, are suitable tools to
describe the causal connections among different nodes.
%
We illustrate in the Fig.~\ref{F_scheme} the schematic representation
of the region causally connected to a given \ceti{}.
%
For the sake of simplicity, we show in the plot the light travel
distance, i.e., the distance traversed by any signal spreading from
the emitter at light speed (for example, any electromagnetic signal).
%
In this scheme, a light pulse would follow a trajectory represented by
a line at 45 degrees from the axes, given the units of time and
space axes are years and light years, respectively.
%
Panel (a) of the Figure represents the region $R$ (shaded polygon)
which is causally connected to the emitter (left vertex).
%
This region develops after an event (hereafter dubbed A-type event) in
which the node acquires the communication capability, or becomes
''active''.
%
The sphere of first contact (SFC) is centred on the emitter,
represented in the Figure by the left angle of the region $R$.
%
This sphere grows until its radius reaches the D$_{\rm max}$ distance, in
which it would be no longer detectable due to the decrease in the
energy per unit area, which falls under the assumed detectability
threshold.
%
Similarly, the surface of last contact (SLC) is another sphere that
grows from an event in which the nodes ceases to possess the
communication capability (D type event), and carries the last signal
produced by the emitter.
%
The region which is causally connected to the emitter is then limited
by these two spheres and therefore has the shape of a filled sphere
(before the D event) or of a spherical shell (after the D event).
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b) of the Figure.
%
In this scheme, the entire lifetime of a \ceti{} can be represented by
a polygon, as shown in the Figure.
%
The polygon is limited in the spatial axis by the maximum distance of
the signal $D_{\rm max}$ and in the time axis by the time span between the
A-type event (concave vertex) and the D-type event (convex vertex).
%
This representation allows to visualize the important events that
result from the presence and communications of two \cetis{}.
%
The points in time where the \cetis{} acquire their communicating
capacity are dubbed ‘’Awakening’’ events ($A_1$ and $A_2$ in the
Figure).
%
Similarly, the instants in time where the \cetis{} lose their
communicating capacity are dubbed ‘’Doomsday’’ events ($D_1$ and $D_2$
in the Figure).
%
The points in space--time where the first contact is produced for each
one of the \cetis{}, are defined as ''Contact'' events, shown as $C_1$
and $C_2$.
%
Finally, the two points where the contact is lost for each one the
\cetis{}, are denominated ''Blackout'' events, presented as $B_1$ and
$B_2$ in the Figure.
%
The receiver node E$_1$ can listen signals from the emitter node
E$_2$, from the first contact event (t=C$_{12}$) until the last
one (t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$.
%
The time intervals for the open communication channel are determined
by the ''C'' and ''B'' type events.
%
It is noteworthy the fact that there is a time delay between the
contact events of the two \cetis{} involved in this analysis, and also
a time delay between the two blackout events.
%
Therefore, the time range when a bidirectional contact is possible
occurs between the maximum time of the contact points and the minimum
time between the blackout points.

The temporal structure that emerges from the experimental setup implies that, as a fundamental property of the simulations, a causal connection can be produced without 
requiring that the two nodes are active at the same time.
%
This property of the system arises as a consequence of the large spatial and temporal scales, where a message is transmitted at the (relatively small) light speed.
%
Although a \ceti{} could be active for a large enough period to transmit a message at large distances, the limited power of the message and the dilution that depends on the squared distance from the source imposes a detectability limit.
%
As a consequence of this limitation and of their finite lifetime, considered as the period between the acquisition and loss of communicating capacity, each \ceti{} will fill a spherical shell region of the galaxy, limited by two concentric spherical surfaces.
%
The leading front, or surface of first contact (SFC) grows from the
central \ceti{} until it reaches the maximum distance $D_{\rm max}$.
%
Following the end of the civilization, there is still a region which is filled with the emitted signals.
%
This approach has been also considered in other statistical models \citep[e.g., ][]{smith_broadcasting_2009, grimaldi_signal_2017, Grimaldi2018}.
%
The trailing front, or surface of last contact (SLC) also grows from the central \ceti{}, with a delay with respect to the SFC equivalent to the lifetime of the node, and produces a spherical shell region.
%
Any other \ceti{} within this region will be aware of the originating node, even if it has disappeared before the time of contact.
%
This region will grow if the surface of first contact has not yet
reached the maximum distance $D_{\rm max}$, and will shrink after a
D-type event until the surface of last contact reaches $D_{\rm max}$, producing as a result the loss of all signal from the central \ceti{}.
%
In our approach, we consider a model galaxy where the width of the disk is negligible with respect to the radius of the disk.
%
In the 2D simulation only the intersection of the communicating spherical shells with the plane of the galaxy is relevant, and produce the corresponding circles or rings for the filled spheres or annular regions of the spherical shells, respectively.
%
The initially growing communicating sphere is shown over space--time diagrams, where time is represented on the horizontal direction, and space is represented in the vertical direction.
%
In the Fig.~\ref{F_scheme} the two emitters in panel (b), E$_1$ and E$_2$, reach each other at different times.
%
The time span for E$_i$ is (A$_i$ , D$_i$) , for $i = {1, 2}$.
%
Emitter $i$ can listen to emiter $j$ between C$_{ij}$ and B$_{ij}$.
%
The type and length of causal contact in both directions depend on the distance and time lag between the awakening events, the maximum distance that a signal can reach and the time period in which each emitter is active.

@@

 \setlength{\tabcolsep}{10pt}
\begin{table*}
\centering
\begin{tabular}{cllll}
\hline
   \multicolumn{2}{l}{independent variables (hypothesis space)}
   &min. value&max. value&Nbins\\
\hline
   $\tau_{a}$ & Mean temporal separation between consecutive awakenings 
	      & 500~a & 40500~a & 21\\ 
	    & & 5$\cdot$10$^5$~a & 10$^6$~a & 51\\ 
   $\tau_{s}$ & Mean lifetime of a \ceti{}
	      & 500~a & 40500~a & 21\\ 
	    & & 5$\cdot$10$^5$~a & 10$^6$~a & 51\\ 
	D$_{\rm max}$ & Maximum reach of a message  & \multicolumn{3}{c}{100 pc, 500 pc, 1 kpc, 5 kpc, 10 kpc} \\
\hline
   \multicolumn{2}{l}{fixed variables} & \multicolumn{2}{c}{assumptions} &value \\
\hline
   & statistical properties of all \cetis{} &equally distributed&&\\
   & Point process for the distribution in time & homogeneous &&\\
   f$_s$ & The scan of the sky & fully efficient&&1\\
   f$_p$ & panspermia or colonization &absent&&0\\
   & shape of the Galactic Habitable Zone & 2--dimensional ring &&\\
	R$_{GHZ}^{\rm min}$   & Inner radius of the GHZ  & \citet{lineweaver_galactic_2004} & & 7~kpc\\
   R$_{GHZ}^{\rm max}$   & Outer radius of the GHZ       & \citet{lineweaver_galactic_2004} & & 9~kpc\\
	t$_{\rm max}$ & Time span of the simulation  & & \multicolumn{2}{r}{5$\cdot$10$^6$~a, 10$^7$~a} \\
    & \multicolumn{3}{l}{number of random realizations for each point in the hypothesis space} & 10, 5, 1 \\
\hline
   \multicolumn{2}{l}{discrete events} &
   \multicolumn{3}{l}{affected variables}\\
\hline
   A event & Awakening: a \ceti{} starts its communication
   capabilities &Number of active \cetis{}\\
   B event & Blackout: the end of the communication channel stops
   &Number causal channels\\
   C event & Contact: a new causal contact is produced &Number of causal channels\\
   D event & Doomsday: a \ceti{} ends its communication capabilities&Number of active \cetis{}\\
\hline
